The Key to a Data Parallel Compiler

### 2015-10-30 2016-pldi/2016-pldi-key_to_data_parallel_compiler.tex
Add some notes and outline
Rewrite the Introduction

\maketitle
\begin{abstract}
By choosing an appropriate representation for the AST, we can implement a compiler for a modern functional 
language using only data-parallel array operations under simple functional composition without the use of 
branching, recursion, or other means of explicit control flow. In this pearl we elucidate the core data structure 
and an operator, called Key, that enable a surprisingly direct implementation of non-trivial transformations. 
The result permits arbitrary computation over sub-trees chosen by parent-child and other inter-node 
relationships in a pure data-parallel fashion without complex control flow. We give direction on the use of this 
method through examples of function lifting and expression flattening.
\end{abstract}

@@

>>>>

\maketitle
\begin{abstract}
TBW
\end{abstract}

%[c]Proposed new outline:
%[c]
%[c]I. 	Generalization on manipulations over Sub-tree computation
%[c]  A.	Key + Encoding → Data parallel, Array-based computation
%[c]  B.	Easy to write/read
%[c]  C.	Uniform across passes
%[c]II.	Case Studies
%[c]  A.	Function Lifting
%[c]  B.	Expression Flattening
%[c]  C.	Lexical Scope Resolution
%[c]  D.	Illustrates the problem space
%[c]III.	Other passes/Issues
%[c]  A.	Discussion of other passes
%[c]  B.	Beyond the Key Operator
%[c]IV.	In practice/Co-dfns Compiler
%[c]V.	Performance/Results
%[c]
%[c]Questions that must be addressed:
%[c]
%[c]1. Why the formalization in APL?
%[c]2. Why parallelize a compiler?
%[c]3. How well does it work/scale/generalize

@@

Compilers are everywhere, but their design and structure does not scale well to modern 
parallel architectures. Traditional compilers are very difficult to parallelize, and the benefits of 
modern architectures is often seen only through the use of distributed build systems which invoke 
the single threaded compilers on multiple ``documents'' at the same time. Future CPU architectures 
promise to be even more data parallel, such as the Xeon Phi and the GPGPU architectures. On these 
systems it is even harder to obtain good performance for compilers written in the traditional fashion.

Fundamentally, a compiler is a translator that operates over some sort of tree and transforms it to 
another tree of an usually radically different shape but equivalent ``semantics.'' These algorithms and 
transformations show up everywhere, from document processing on the Web, XML-based publishing 
workflows, to the programming language compilers with which most developers are intimately familar. 
These transformations differ from other widely studied parallel graph problems in that they tend to greatly 
alter the fundamental structure of the graph/tree, unlike, say, breadth-first search or single-source 
shortest path algorithms. This rapidly shifting graph structure requires different approaches to 
achieving good parallelism.

There has been some interesting use of GPUs to accelerate certain compiler analyses and passes. 
Unfortunately, attempts to parallelize compilers usually takes the form of a great deal of work spent to 
highly optimize a single compiler pass. The results are often fragile and require a great deal of effort 
to maintain and keep in line with the rest of the compiler. This integration problem makes it difficult to 
scale the "ad hoc" parallelization effort of individual compiler passes to a complete system. This is 
particularly true in situations where the compiler passes may change rapidly, such as with XSLT or 
document publishing, where it is very common for the document preparer to write or tweak transformation 
passes to suit the specific class of documents being handled. In these cases, the expertise and time 
involved in parallelizing such transformations makes the ad hoc approach mostly intractable.

Furthermore, compilers are increasingly becoming a part of a "live" and interactive feedback loop. 
Just in time compilation and on the fly static compilation are both used to provide custom binaries tuned 
to specific tasks on the fly. These techniques are used in web services and large scale databases, where 
in-memory databases may construct functions on the fly to execute. There is great interest in moving 
such systems on to the GPU or other highly data-parallel architectures, but it would significantly hamper 
the performance of the system if the compilation phase of query execution were required to occur on the 
CPU, especially at scale where there may be many more GPU accelerators than CPU resources.

We propose a language-driven approach to compiler construction that provides a framework for the 
development of arbitrary compiler transformations that are data-parallel by construction. These 
techniques intentionally optimize for the data-parallel case, which permits a wide degree of platform 
independent performance compared to the traditional approach to compiler design, which requires 
significant effort and possibly signficiant transformation of the source code to achieve good parallel 
performance on high-parallel architectures.

We use a mathematically oriented, array based language that is accessible and popular with information 
scientists. We restrict this language to permit only the composition of data-flow oriented array based 
programs whose core primitives are all data-parallel by definition. The permitted compositions of these 
primitives ensures that the resulting programs are both analyzable and parallel by construction, permitting 
a wide array of whole program and transformation specific optimizations that have been long studied. 
The notation is intentially close to mathematical notation, and as such, the resulting compiler passes 
more closely resemble the compact formal descriptions of the transformations rather than complex 
recursive algorithms for transforming trees.

At the core of such techniques, we must have some method for transforming trees in a controlled and 
predictable manner. We build on the Nanopass paradigm for compiler construction, wherein a compiler is 
developed as a series of small, functional tree transformations (passes) that have an input AST language 
and an output AST language. Through progressive refinement of the tree, the composition of these 
passes results in the complete compiler. This is a highly functional approach, with very little mutation. 
Fundamentally, each pass must identify regions of the tree upon which it works, do some sort of computation 
on that sub-tree that has been matched (often through pattern matching on the nodes) and reassemble 
the tree in some new order that matches the desired structure. In other words, a compiler nanopass 
can be thought of as an arbitrary computation over a given selection of sub-trees in an AST, where 
this computation could move, relocate, or change the sub-tree.

The key to our language-drive approach is a set of techniques that enable a data-parallel expression of 
these arbitrary transformations. By using a higher-order function called Key, together with basic array 
programming techniques and an encoding for the tree connections called a Node Coordinate matrix, we can 
identify sub-trees and select them for computation by their parent-child relationships in a data-parallel 
fashion, and then compute on these sub-trees using traditional array programming techniques.

We have constructed a complete compiler for a superset of our language called Co-dfns. This compiler is 
part of a commercial effort to develop platform independent performance for array programs. The 
compiler is developed completely in a data-parallel by construction as described in this paper, and uses 
only traditional and well known array primitives together with function composition to form the entire 
core compiler passes and generator. As a result, the compiler takes the form of a pure data-flow graph 
without mutation and has an extremely uniform and concise definition. It uses no recursion, pattern 
matching, branching, or other forms of explicit looping. The compiler, together with its runtime libraries 
and parser, comes in under 1000 lines of code.

In this paper we fully elucidate three compiler passes from the Co-dfns compiler to demonstrate the 
techniques used and we provide further analysis of how these techniques are used in the other compiler 
passes. We further provide a comparative analysis of these passes against their counterparts implemented 
using the Nanopass compiler framework, comparing performance both in terms of programmer effort and 
portability/maintainability, as well as runtime performance.

\paragraph{Contributions}

\begin{itemize}[noitemsep]
\item We describe a method of computing over arbitrary sub-trees selected by their parent-child 
relationships in a data-parallel manner using the Key operator and Node Coordinates.
\item We situate this technique into a broader, language-driven compiler construction strategy 
that enables the development of parallel by construction compilers that exhibit desirable programmability 
features in terms of code size, complexity, portability, and maintainability. 
\item We provide analysis of these techniques as used in a commercial compiler project called Co-dfns and 
report on the results.
\item We compare these techniques to the Nanopass approach to compiler design in terms of programmer 
productivity and compiler runtime performance. 
\item We enumerate a set of programming techniques, patterns, and strategies that complement the 
Key operator and Node Coordinates in the construction of tree transformations in general and give some 
intuitions on the use of these techniques on arbitrary transformations.
\end{itemize}
%[cf]
%[of]:Introduction
\section{Introduction (Old)}
